Relational Parametricity for Algebraically Indexed Types

### 2012-06-08 paper/section-general-framework.tex
Latest changes

We next present the semantics of types in
\autoref{sec:semantics-algebraically-indexed-types}. Types are first
interpreted without reference to the indexing terms: the
\emph{index-erasure} interpretation. The index-erasure interpretation
provides an ``underlying'' semantics to each type (and term), which we
think of as an abstraction of the actual operational behaviour of a
term. By erasing all the index terms, we can immediately see that the
run-time behaviour of programs cannot depend on index terms. We then
define the \emph{relational interpretation}, which interprets every
type as a binary relation on its index-erasure interpretation. The
relational interpretation acts as ``commentary'' on the underlying
interpretation, constraining the index-erasure interpretations that
are allowed. In particular, we use the relational interpretation to
enforce the invariance properties that we indentified in
\autoref{sec:motivating-examples}.

@@

We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given a relational environment
$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$. \fixme{Check this is the right way round}

@@

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.~ subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we may have different sets of relational
environments that enforce different invariants. For the general
framework, we assume that, for each index context $\Delta$, we are
given a set of relation environments $\relEnv{E}(\Delta)$. Note that
this assignment of sets of relation environments to index contexts is
not assumed to be compositional: it is not necessarily the case that
the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
defined in terms of $\relEnv{E}(\Delta_1)$ and
$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow
  \Delta$, then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
  : \Delta' \Rightarrow \Delta$, such that the outer edges of the
  following diagram commute, for all primitive types $\tyPrimNm{X} \in
  \mathit{PrimType}$:
  %FIXME: check that these compositions are the right way round
  %FIXME: re-do this as a set of equations
  %FIXME: give a better English description of why this is needed
  \begin{displaymath}
    \xymatrix{
      {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
      \\
      {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
      \\
      &
      &
      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
  two triangles in the bottom right of the diagram commute.
  % FIXME: mention that we already know that the square commutes
\end{enumerate}

@@

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeqsubst-relational}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho'}$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:ctxtsubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}{\rho''}$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:ctxtsubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}

>>>>

We next present the semantics of types in
\autoref{sec:semantics-algebraically-indexed-types}. Types are first
interpreted without reference to the indexing terms: the
\emph{index-erasure} interpretation. The index-erasure interpretation
provides an ``underlying'' semantics to each type (and term), which we
think of as an abstraction of the actual operational behaviour of a
term. By erasing all the index terms, we can immediately see that the
run-time behaviour of programs cannot depend on index terms. We then
define the \emph{relational interpretation}, which interprets every
type as a binary relation on its index-erasure interpretation. The
relational interpretation acts as ``commentary'' on the underlying
interpretation, constraining the index-erasure interpretations that
are allowed. In particular, we use the relational interpretation to
enforce the invariance properties that we identified in
\autoref{sec:motivating-examples}.

@@

We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given a relational environment
$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.

@@

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.~ subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we may have different sets of relational
environments that enforce different invariants. For the general
framework, we assume that, for each index context $\Delta$, we are
given a set of relation environments $\relEnv{E}(\Delta)$. Note that
this assignment of sets of relation environments to index contexts is
not assumed to be compositional: it is not necessarily the case that
the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
defined in terms of $\relEnv{E}(\Delta_1)$ and
$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow
  \Delta$, then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
  : \Delta' \Rightarrow \Delta$, such that $\rho_1 \circ
  \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists a $\rho
  \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho \circ
  \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ \pi_{i\mathord:s'} =
  \rho_2$.

  % such that the outer edges of the
  % following diagram commute, for all primitive types $\tyPrimNm{X} \in
  % \mathit{PrimType}$:
  % %FIXME: check that these compositions are the right way round
  % %FIXME: re-do this as a set of equations
  % %FIXME: give a better English description of why this is needed
  % \begin{displaymath}
  %   \xymatrix{
  %     {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
  %     &
  %     {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
  %     \\
  %     {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
  %     &
  %     {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
  %     \\
  %     &
  %     &
  %     {\Rel(\tyPrimSem{\tyPrimNm{X}})}
  %   }
  % \end{displaymath}
  % Then there exists a relation environment $\rho \in
  % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
  % two triangles in the bottom right of the diagram commute.
  % FIXME: mention that we already know that the square commutes
\end{enumerate}

@@
